### 2018-01-24 includes/abstract.tex
Fleshing out sections and revising a bunch of content further.

    In this paper we present \fuzzer{}: an \smtlib{} problem fuzzer and generator for string solvers. We provide a rich repository of problems we generated with \fuzzer{} in \smt{} format. We also analyse \numSolvers{} mature string solvers: \theSolvers{}. We use \fuzzer{} to craft inputs that elicit poor performance in \us{} and \cvc{}, and analyse the causes of the performance degradations. We then describe an algorithm to mitigate the performance degradation in \us{}. Finally, we describe several bugs in \us{} that were exposed by \fuzzer{}.

>>>>

    In this paper we introduce \fuzzer{}: a modular \smt{} problem transformer and generator for string solvers. We supply a repository of problems we generated with \fuzzer{} in \smtfull{} format, and demonstrate its usefulness by showing defects it helped find and fix in a mature solver: \us{}. We also test the solvers \theSolvers{} with suites of \fuzzer{} problems that elicit poor performance in some of the solvers, but not in others. We briefly analyse our experimental results, and identify the causes of the performance degradations in \us{}.

### 2018-01-24 includes/experiment.tex
Updating graphs, adding some docs to Makefile, and putting graph captions under graphs..
Fleshing out sections and revising a bunch of content further.

\section{Experimental Data}

    We generated several problem suites with \fuzzer{} that made one solver perform poorly, but not others. The graphs in the figures below show the performance results. All experiments were run in series on the same Linux computer, with a timeout of 15 seconds.

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \caption{Performance on concats-balanced}
            \label{fig:concats-balanced}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \caption{Performance on concats-small}
            \label{fig:concats-small}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \caption{Performance on concats-extracts-small}
            \label{fig:concats-extracts-small}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \caption{Performance on different-prefix}
            \label{fig:different-prefix}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}
        \end{subfigure}
        \caption{Results on selected \fuzzer{} problems}
        \label{fig:graphs}
    \end{figure}

\section{Analyses}

    In this section, we analyse the experimental results from the previous section. Although it is satisfying to find a class of problems that makes only one solver behave a certain way, it is only practically useful when the behaviour's cause is also understood.

    \todo{Maybe organise these by solver, instead of by problem?}

    \subsection{\cHard{}}

        The \cvc{} trace file suggests that it performs the following operations:

        \begin{enumerate}
            \item Loads all facts into its database.
            \item Flattens binary concatenations into n-ary concatenations.
            \item Computes equivalence classes for strings and integers.
        \end{enumerate}

        At this stage it discovers that there is a conflict in one of its equivalence classes, as shown in Figure~\ref{fig:cvc4-conflict}. After discovering the conflict, the rest of the trace is filled with apparently the same operation repeated hundreds of thousands of times. This operation dominates the content of the trace. After this expensive operation is finished, \cvc{} appears to recall the earlier conflict and returns \texttt{unsat}.

        \begin{figure}[h]
            {\scriptsize\begin{verbatim}
CONFLICT: Eq engine conflict : (and
(= "woof" (str.++ "woof" lsym_5 ... lsym_5))
(= "" lsym_5)
(= "" var3) ... (= "" var30)
(= (str.++ "woof" var3 ... var30) (str.++ "meow" var43 ... var70))
(= "" var43) ... (= "" var70)
(= "meow" (str.++ "meow" lsym_5 ... lsym_5)))\end{verbatim}}
            \caption{Evidence of \cvc{} detecting conflict}
            \label{fig:cvc4-conflict}
        \end{figure}

        \us{} behaves differently. It follows a similar process as \cvc{}, except it does not flatten concatenations. In contrast to \cvc{}, \us{} immediately returns \texttt{unsat} as soon as it verifies consistency of equivalence classes as it constructs them, as shown in Figure~\ref{fig:z3str3-conflict}.

        \begin{figure}[h]
            {\scriptsize\begin{verbatim}
-------- [str] new_eq_check ../src/smt/theory_str.cpp:1864 ---------
checking whether (str.++ "woof" var2) and (str.++ "meow" var42) can be equal
------------------------------------------------
-------- [str] new_eq_check ../src/smt/theory_str.cpp:1867 ---------
inconsistency detected: (str.++ "woof" var2) cannot be equal to (str.++ "meow" var42)
------------------------------------------------
-------- [str] assert_axiom ../src/smt/theory_str.cpp:179 ---------
asserting (not (= (str.++ "woof" var2) (str.++ "meow" var42)))
------------------------------------------------\end{verbatim}}
            \caption{Evidence of \us{} detecting conflict}
            \label{fig:z3str3-conflict}
        \end{figure}

        We were not equipped well enough to debug this behaviour in \cvc{}, but we intend this result to be useful to its authors in improving it.

    \subsection{\zHard{}}

        Both solvers exhibit similar patterns in their trace files. They differ in their strategies for assigning possible satisfying assignments.

        \todo{Explain why \us{} performs poorly.}

>>>>

\section{Experimental Data}
\label{sec:data}

    We generated several problem suites with \fuzzer{} that made one solver perform poorly, but not others. These suites are \theSuites{}. Figure~\ref{fig:cvc-hard} shows the suites that were uniquely difficult for \cvc{}. Figure~\ref{fig:z3str3-hard} shows the suites that were uniquely difficult for \us{}. All experiments were run in series on the same Linux computer, with a timeout of 15 seconds.

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-extracts-small.png}
            \label{fig:concats-extracts-small}
            \caption{Performance on concats-extracts-small}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/different-prefix.png}
            \label{fig:different-prefix}
            \caption{Performance on different-prefix}
        \end{subfigure}
        \caption{Problems hard for \cvc{}}
        \label{fig:cvc-hard}
    \end{figure}

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-balanced.png}
            \label{fig:concats-balanced}
            \caption{Performance on concats-balanced}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \includegraphics[width=\textwidth]{data/graphs/concats-small.png}
            \label{fig:concats-small}
            \caption{Performance on concats-small}
        \end{subfigure}
        \caption{Problems hard for \us{}}
        \label{fig:z3str3-hard}
    \end{figure}

\section{Analyses}
\label{sec:analysis}

    In this section, we analyse our experimental results. We discuss the causes of \us{}'s poor performance on the two problem suites: \textit{concats-balanced} and \textit{concats-big}. We were not equipped well enough to debug the behaviours of the other solvers, but we provide our data in hopes that they will be useful to the solvers' authors.

    \subsection{\us{} on concats-balanced}

        \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

% X = "solution"
% X = (((A.B).(C.D)).((E.F).(G.H))).
%     (((I.J).(K.L)).((M.N).(O.P)))

    \subsection{\us{} on concats-big}

        \todo{Dmitry with help: Explain why \us{} performs poorly on this suite.}

% X = "solution"
% X = (A.(B.(C.(D.(E.(F.(G.H)))))))

### 2018-01-24 includes/introduction.tex
Addressing Federico's feedback. Adding Makefile target to re-gather data.
Fleshing out sections and revising a bunch of content further.

    Many SMT solvers that support the theory of strings have been developed over the past decade, and several have matured to the point of being performant and useful in practice. \theSolvers{}\cite{cvc4}\cite{z3str3} are \numSolvers{} such solvers. Moreover a standard input language for problem instances, \smt{}\cite{smt}, has been developed and many solvers support it. To validate their solvers, many authors have relied on hand-crafted input suites or real-world examples from industrial applications \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests}. Indeed several suites of industrial and hand-crafted problems have been compiled \cite{kaluza} \cite{kausler} to both validate solvers and benchmark their performance. However, though these suites help debug solvers, they do not age well. As solvers continue to grow, fixed benchmarks leave more and more functionality untested.

    We introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We show its effectiveness by presenting defects we were able to find in leading solvers.

@@

    \begin{enumerate}
        \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smt{} string and regex problems. We briefly document its components, modular architecture, and usage patterns to demonstrate its usefulness as an exploratory tool for validating not only solver implementations, but also algorithmic assumptions.
        \item \textbf{A suite of \smt{} problems}: We present a suite of \smt{} string and regex problems we generated with \fuzzer{}. We justify the existence of each suite by a motivating example from the development of \us{}. We reveal specific defects that this suite helped uncover and fix, as well as specific conjectures that it helped validate or disprove.
        \item \textbf{Experimental Results}: We present problem classes we found that made one solver perform poorly, but not another. We show differences in solver behaviour on the suites \cHard{} and \zHard{}.
        \item \textbf{Analysis}: We analyse our experimental results by inspecting the execution traces of the solvers on the given problems. Where possible, we explain our observations. For \cvc{} we draw superficial conclusions based solely on the execution traces, and invite its authors to join us in analysing them. For \us{} we pinpoint algorithmic shortcomings that cause the poor performance.
    \end{enumerate}

>>>>

    Many SMT solvers that support the theory of strings have been developed over the past decade, and several have matured to the point of being performant and useful in practice. \theSolvers{}\cite{cvc4}\cite{z3str3} are \numSolvers{} such solvers. Moreover a standard input language for problem instances, \smtfull{}\cite{smt}, has been developed and widely adopted. To validate their solvers, many authors have relied on hand-crafted input suites or real-world examples from industrial applications \cite{cvc4-tests} \cite{z3str3-tests} \cite{z3str2-tests}. Indeed several suites of industrial and hand-crafted problems have been compiled \cite{kaluza} \cite{kausler} to both validate solvers and benchmark their performance. However, though these suites help debug solvers, they do not age well. As solvers continue to grow, fixed benchmarks leave more and more functionality untested.

    We introduce a modular fuzzer for string solvers, \fuzzer{}, and describe its utility as an exploratory validation tool. We show its effectiveness by presenting defects and limitations we were able to find in leading solvers.

@@

        \begin{enumerate}
            \item \textbf{The \fuzzer{} tool}: We describe a modular fuzzer which can transform and generate \smtfull{} string and regex problems. We briefly document its components and modular architecture. We provide some motivating usage patterns to demonstrate its usefulness as an exploratory tool for validating solvers. We additionally detail its regex-generating and transforming capabilities.
            \item \textbf{A repository of \smtfull{} problems}: We present a repository of \smtfull{} string and regex problem suites we generated with \fuzzer{}. We demonstrate the usefulness of these suites by revealing specific defects they helped uncover and fix in \us{}.
            \item \textbf{Experimental Results}: We present experimental runs of \theSolvers{} on the \fuzzer{} suites: \theSuites{}. We picked these suites because they uniquely made some solvers perform poorly, but not others.
            \item \textbf{Analysis}: We analyse our experimental results by inspecting the execution traces of the solvers on the given problems. For \us{} we pinpoint algorithmic limitations that cause the poor performance. For other solvers, we draw superficial conclusions based solely on the execution traces, and invite their authors to conduct further analysis.
        \end{enumerate}

### 2018-01-24 includes/problems.tex
Fleshing out sections and revising a bunch of content further.

\section{Problem Suites}

    We generated several suites of string and regex problems using \fuzzer{}. These problems are available at \problemRepo{}. Below, we describe the \texttt{generated} suite that we generated using \generator{}:

    \todo{Fill in descriptions.}

    \begin{tabular}{|c|c|}
        \hline
        \textbf{Name}    & \textbf{Description} \\ \hline
        Lengths          & \\ \hline
        Lengths-Concats  & \\ \hline
        Concats          & \\ \hline
        Concats-Extracts & \\ \hline
        Overlaps         & \\ \hline
        Regex            & \\ \hline
        Many-Regexes     & \\ \hline
        Regex-Deep       & \\ \hline
        Regex-Pair       & \\ \hline
        Regex-Lengths    & \\ \hline
    \end{tabular}
    \hfill \\ \\

@@

    \todo{Fill in descriptions.}

    \begin{tabular}{|c|c|}
        \hline
        \textbf{Name} & \textbf{Description} \\ \hline
        Amazon        & \\ \hline
        Regex         & \\ \hline
    \end{tabular}

    \subsection{Results}

@@

        \todo{Describe the length lookup bug and length-testing bug, giving the before/after commits.}

    \subsection{Patterns}

        While the aforementioned suites are useful in analysing a solver, we also used randomised \fuzzer{} runs for exploratory debugging. For example the following usage can feed random problems to \cvc{} until it returns an error:

        {\scriptsize\begin{verbatim}
while stringfuzzg -r random | tee problem.smt25 | cvc4 --lang smt2.5 --tlimit=5000; do
    sleep 0
done\end{verbatim}}

        Some \unix{} scripting can yield other interesting results. For example the following script finds a regex problem on which \us{} times out after 5 seconds:

        {\scriptsize\begin{verbatim}while true; do
    stringfuzzg -r regex -c -a increasing -r 2 -t 5 -d 3 > problem.smt25
    result=`z3str3 -T:5 problem.smt25`
    echo $result
    if [ "$result" == "timeout" ]; then
        cat hard-problem.smt2
        break
    fi
done\end{verbatim}}

>>>>

\section{Problem Suites}
\label{sec:problems}

    We generated several suites of string and regex problems using \fuzzer{}. These problems are available at \problemRepo{}. Below, we describe the \texttt{generated} suite that we generated using \generator{}: \\

    \begin{tabular}{|l|l|}
        \hline
        \textbf{Name}    & \textbf{Problems have ...} \\ \hline
        Concats          & A single, deep tree of concats. \\ \hline
        Concats-Extracts & A single concat tree, and random character extractions. \\ \hline
        Lengths          & A single, large length constraint on a variable. \\ \hline
        Lengths-Concats  & Variables and their concats, each with fixed lengths. \\ \hline
        Overlaps         & A formula in the form A.X = X.B. \\ \hline
        Regex            & A test for regex membership of arbitrary complexity. \\ \hline
        Many-Regexes     & A test for membership in many random regexes. \\ \hline
        Regex-Deep       & A test for membership in a regex with many nested operators. \\ \hline
        Regex-Pair       & A test for membership in one regex, but not another. \\ \hline
        Regex-Lengths    & A test for membership in a regex, and a length constraint. \\ \hline
    \end{tabular}
    \hfill \\

@@

    \todo{Federico: fill in descriptions.}

    \begin{tabular}{|l|l|}
        \hline
        \textbf{Name} & \textbf{Description} \\ \hline
        Amazon        & \\ \hline
        Regex         & \\ \hline
    \end{tabular}

    \subsection{Usefulness to \us{}}

@@

        The first bug we found caused \us{} to perform poorly when the equivalence class of arithmetic terms is very large (e.g. where many strings are equal to the empty string, and thus have length 0). The slowdown was caused by a loop over the equivalence class of these terms to search for an integer constant. The fix was to check the root of the equivalence class for the integer constant, as Z3 makes "interpreted terms" (i.e. constants) the root of equivalence classes whenever possible. This bug was last observed in commit 6308636, and was fixed in commit 3865c45.

        The second bug we found caused \us{} to perform poorly when a variable had a large length constraint. The slowdown was caused by the solver performing a linear search for a satisfying length when producing a model for the variable. This bug was an inconsistency between the implementation and the behaviour described in the \us{} paper \cite{z3str3}. This bug was last observed in commit 66bc68f, and was fixed in commit 7b536e9.

        \todo{Dmitry: maybe show graphs with before and after?}

### 2018-01-24 includes/stringfuzz.tex
Adding some TODOs.
Fleshing out sections and revising a bunch of content further.

\section{\fuzzer{}}

    This section describes the fuzzer \fuzzer{}, which generates new \smt{} problems and transforms existing ones. \fuzzer{} is implemented as a Python package, and comes with several executables to generate, transform, and analyse \smt{} string problems.

@@

    The \fuzzer{} code is organised to be easily extended: all generators and transformers are independent modules. Its components are also implemented as \unix{} ``filters'', to enable easy integration with other tools (including themselves). For example, the outputs of generators can be piped directly into transformers, and transformers can be chained to produce a tuned stream of challenging inputs to a solver.

@@

        \item[\texttt{stringfuzzg}] \hfill \\
            This tool generates \smt{} problems. It supports several generators and options that specify its output. The generators it implements are: \\

            \todo{Fill in descriptions.}

            \begin{tabular}{|c|c|}
                \hline
                \textbf{Name} & \textbf{Description} \\ \hline
                concats       & \\ \hline
                lengths       & \\ \hline
                overlaps      & \\ \hline
                regex         & \\ \hline
                random-text   & \\ \hline
            \end{tabular}

@@

        \item[\texttt{stringfuzzx}] \hfill \\
            This tool transforms \smt{} problems.. It supports several transformers and options that specify its output and input. The transformers it implements are: \\

            \todo{Fill in descriptions.}

            \begin{tabular}{|c|c|}
                \hline
                \textbf{Name} & \textbf{Description} \\ \hline
                fuzz          & \\ \hline
                graft         & \\ \hline
                multiply      & \\ \hline
                nop           & \\ \hline
                reverse       & \\ \hline
                rotate        & \\ \hline
                translate     & \\ \hline
                unprintable   & \\ \hline
            \end{tabular}

@@

            \todo{Add details about merging strategies.}

@@

>>>>

\section{\fuzzer{}}
\label{sec:fuzzer}

    This section describes the fuzzer \fuzzer{}, which generates new \smtfull{} problems and transforms existing ones. \fuzzer{} is implemented as a Python package, and comes with several executables to generate, transform, and analyse \smtfull{} string problems.

@@

    The \fuzzer{} code is organised to be easily extended: all generators and transformers are independent Python modules. Its components are also implemented as \unix{} ``filters'', to enable easy integration with other tools (including themselves). For example, the outputs of generators can be piped directly into transformers, and transformers can be chained to produce a tuned stream of challenging inputs to a solver.

@@

        \item[\generator{}] \hfill \\
            This tool generates \smt{} problems. It supports several generators and options that specify its output. The generators it implements are: \\

            \begin{tabular}{|l|l|}
                \hline
                \textbf{Name} & \textbf{Generated Problems have ...} \\ \hline
                concats       & Long concats and optional random extracts. \\ \hline
                lengths       & Many variables (and their concats) with length constraints. \\ \hline
                overlaps      & An expression of the form A.X = X.B. \\ \hline
                equality      & An equality among concats, each with variables or constants. \\ \hline
                regex         & Regexes of variying complexity. \\ \hline
                random-text   & Random, likely syntactically invalid text. \\ \hline
                random-ast    & Totally random, but semantically valid text. \\ \hline
            \end{tabular}

@@

        \item[\transformer{}] \hfill \\

            This tool transforms \smt{} problems.. It supports several transformers and options that specify its output and input. The transformers it implements are: \\

            \todo{Dmitry: maybe add notes about SAT preservation?}

            \begin{tabular}{|l|l|}
                \hline
                \textbf{Name} & \textbf{The transformer ...} \\ \hline
                fuzz          & Randomly replaces literals and operators with similar ones. \\ \hline
                graft         & For each sort, swaps a random non-leaf node with a leaf node. \\ \hline
                multiply      & Multiplies each number by N, and repeats each character N times. \\ \hline
                nop           & Does nothing. \\ \hline
                reverse       & Reverses all string literals and concat arguments. \\ \hline
                rotate        & Performs a random concat tree rotation. \\ \hline
                translate     & Maps every character X to a different character Y. \\ \hline
                unprintable   & Replaces all characters with unprintable ones. \\ \hline
            \end{tabular}

@@

            \todo{Federico: describe \texttt{stringmerge}.}

@@

    \subsection{Regex Fuzzing Capabilities}

        \fuzzer{} can generate and transform problems with regular expression constratints. The command \texttt{stringfuzzg regex} invokes the regex problem generator, and has several options to control its output. It generates a problem in the form\footnote{Variable declarations omitted for brevity.}:

        \begin{align*}
            \langle problem \rangle  & ::= \langle membership \rangle \; \langle membership \rangle * \\
            \langle membership \rangle & ::= \texttt{(assert (str.in.re X}\; R\; \texttt{))}
        \end{align*}

        where $R \in RegEx$. More simply, the problem is a set of one or more regex constraints on a single variable.

        \todo{Dmitry: further describe the properties of $R$.}

    \subsection{Patterns}

        \fuzzer{} can be used to generate static benchmarks as described in Section~\ref{sec:problems}. It cna also be used in a randomised fashion for exploratory debugging. As a motivating example, the following usage can feed random problems to \cvc{} until it returns an error:

        {\scriptsize\begin{verbatim}
while stringfuzzg -r random --num-terms 1000 | tee problem.smt25 | cvc4 --lang smt2.5 --tlimit=5000; do
    sleep 0
done\end{verbatim}}

        Some \unix{} scripting can yield other interesting results. For example the following script finds a regex problem on which \us{} times out after 5 seconds:

        {\scriptsize\begin{verbatim}while true; do
    stringfuzzg -r regex -c -a increasing -r 2 -t 5 -d 3 > problem.smt25
    result=`z3str3 -T:5 problem.smt25`
    echo $result
    if [ "$result" == "timeout" ]; then
        cat hard-problem.smt2
        break
    fi
done\end{verbatim}}
